The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-08-31 adaptation/article/article.tex
First draft. Before letting it sit

While these general models of adaptation explain the most common phenomena observed in experiments, a number of known phenomena remain outside of their scope. For example, it is known that adaptation rate is reduced in situations where the environment is unstable and unpredictable \citep{Herzfeld_memory_2014}, or situations in which errors are small \citep{Marko_Sensitivity_2012} and slowly introduced \citep{Huang_Persistence_2009}. Learning new adaptations has been found to depend on the history of adaptations learned \citep{Vaswani_Decay_2013}, as well as the way in which those adaptations were presented [CITATION NEEDED].

@@

\subsection{The model}
The model we present has three main components: (1) context-inference, (2) motor adaptation and (3) action selection. Each of these components informs the ones that follow: motor adaptation is informed by context inference, and action selection is informed directly by both motor adaptation and by context inference.

We now briefly describe the three components of the model separately. For clarity, we start with the second component (i.e. motor adaptation), in order to introduce the terminology that we use throughout this work.

@@

\subsection{Context inference}
In our model, the idea of context is quite general. The context comprises the relevant elements of the environment (e.g. the strength and direction of the wind when walking) and of the task at hand (immediate and long-term goals, reward structure, etc). Previous models \cite[e.g.][]{Wolpert_Multiple_1998,Imamizu_Neural_2008,Oh_Minimizing_2019} have focused on the elements of the context which are directly related to the internal models: given a past action, the different internal models (e.g. two forward models $f_B$ and $f_H$) make different predictions for the outcome of that action; the model that best predicts that action is deemed to be the ``correct'' model, i.e. the model that best represents the context.

@@

\subsection{Action selection}
Action selection is affected by the current context via the context-dependent forward model that is used. In our model, action selection is made by building a distribution over available actions which is a weighted sum of the distributions given by the existing forward models, where the weights are obtained by the context inference component of the model. In other words, a distribution is created as follows:
\begin{equation}
p(c_t) \propto \displaystyle\sum_{\zeta \in \Phi}q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) p(c_t | \zeta)
\end{equation}
where $p(c_t | \zeta)$ corresponds to \eref{eqn:forward-model}. From this distribution, the current action (motor command $c_t$) can be sampled and carried out.

\section{Experimental results}
In this section, we go through a number of experimentally-observed phenomena and show that context inference, as done in our model, provides a parsimonious explanation for all of them.

For clarity, we introduce some terminology before discussing these phenomena. As an example, we will use a typical motor adaptation task in which participants have to make forward-backward reaching movements while holding the handle of a mechanical arm that exerts a force. Depending on the trial, the arm might exert a curl force in a clockwise or counter-clockwise direction, or no force at all. Let us define the baseline context O as that in which the mechanical arm exerts no force. Context A can be defined as that in which the arm exerts the clockwise curl force and context B counter-clockwise. Additionally, abusing notation, it can be said that $B = (-A)$, as the forces point in opposite directions. We can also talk of contexts such as A and A/2, which means that both have the same direction of the adaptation (e.g. clockwise), but the second one has half the magnitude. Finally, error-clamp trials, in which the mechanical arm forces the participant to make straight-line movements, are represented with the letter E.

@@

\subsection{Savings}
Savings refers to the ability of humans and other animals to remember a previously-learned adaptation and apply it without having to re-learn it. Savings is almost universally observed in human participants \cite{Brashers-Krug_Consolidation_1996,Shadmehr_Functional_1997,Medina_Mechanism_2001,Smith_Interacting_2006,Zarahn_Explaining_2008}. In an O-A-O-A experiment, for example, savings would be seen in the second A block in the form of a much higher adaptation rate higher than that observed during the first A block.

In this section, we discuss savings, and the related phenomenon of deadaptation, in terms of the switching between forward models. This has been modeled before \citep[e.g.][]{Wolpert_Multiple_1998,Oh_Minimizing_2019}, but we show that expanding switching into context inference explains facets of savings that these previous models do not.

@@

To our knowledge, there are no experiments in which no contextual information is available. However, many experiments \citep[e.g.][]{Huang_Persistence_2009,Brennan_Decay_2015,Smith_Interacting_2006} contain blocks of error-clamp trials in which participants have no way of identifying the current context. These blocks could be considered uncued, however, as we discuss in the following sections, contextual information is still available to participants.

In this section, we focus on cued and partially-cued experiments and leave error-clamp trials to the following sections.

@@

\begin{figure}
\centering
\includegraphics{./figures/figure_1.png}
\caption{Savings and de-adaptation. Data from our simulations (left column) compared to data from \cite{Kim_Neural_2015} and \cite{Oh_Minimizing_2019} (right column). In all panels, the blue line represents the adaptation displayed by the agent as a function of trial number. The black lines represent the optimal adaptation, i.e. the size of the visuomotor rotation during the task. (A) Experiment by \cite{Kim_Neural_2015} with two rotations (-40 and 40 degrees), in addition to baseline. The switches between contexts were visually cued and therefore switches are immediate, as can be seen in both simulations (left) and the experimental data from \cite{Kim_Neural_2015} (right). (B) An experiment in which the agent must adapt to a 20 degree visuomotor rotation during shooting movements. It can be seen in both the simulations (left) and experimental data from \cite{Oh_Minimizing_2019} (right) that after context switches (e.g. at trial 120), the agent returns to the adaptation of the previous block, with one or two trials of delay. (C) Same experiment as (B) but  with a 10 degree adaptation. In this case, savings are much slower.} \label{fig:oh-2019}
\end{figure}

\cite{Oh_Minimizing_2019} performed a similar experiment in which contexts were not visually cued. In addition, the authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. \cite{Oh_Minimizing_2019} immediate switching in the first experiment, while switching was much slower in the second experiment. As can be seen on the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large, and slow, noisy switches when it is low. This difference is explained by our model in terms of the size of the adaptation: the smaller the adaptation is, the more difficult it is for the agent to notice when the context has changed. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

@@

\section{The effects on the rate of adaptation}
In our model, adaptation is gated by the uncertainty on the current context. More specifically, a prediction error observed at trial $t$ will effect motor adaptation with a magnitude proportional to the size of the error, where the proportionality constant is related to the uncertainty over the context in which the error was observed: the higher the uncertainty, the lower the size of the adaptation.

The most direct evidence for the effects of context inference on the rate of adaptation come from the experiments by \cite{Herzfeld_memory_2014}. The authors showed that the volatility of the environment, expressed in terms of unpredictable, stochastic transitions between A and -A (and vice versa) change the speed at which adaptation occurs.

\cite{Herzfeld_memory_2014} used an experimental paradigm in which the force that a mechanical arm exerts on the participant's hand can change from trial to trial between O, A and -A. They divided participants into three groups for which the context transition from one trial to the next was very uncommon, somewhat common and almost every trial. Importantly, these transitions were uncued, and the participant could only infer that a transition had occurred after a motor error was observed.

@@

\begin{figure}
\centering
\includegraphics{./figures/figure_2.png}
\caption{Learning under unstable and unpredictable contexts. 30 trials were simulated in the task defined by \cite{Herzfeld_memory_2014}, repeated 300 times for better statistics. Two conditions were used, following the naming convention from \cite{Herzfeld_memory_2014}: $z=0.5$ is an unpredictable condition, where the context has a 0.5 probability of switching from trial to trial. $z=0.9$ is a stable condition, where the context has a 0.9 probability of staying the same from trial to trial. (A) Adaptation as a function of trial number, averaged across all runs. (B) Error sensitivity, binned into bins of 5 trials, averaged across all runs. The error is in arbitrary units, causing sensitivity to error (y axis) to also be in arbitrary units. Shaded areas are SEM.}
\label{fig:herzfeld-2014}
\end{figure}

To account for volatility, our model includes a term in context inference which assumes a transition from the previous context to all other possible contexts (see Methods). Effectively, this transition term gives the model uncertainty on the current context, regardless of observations. The higher the volatility, the higher the uncertainty. This higher uncertainty is what lowers the observed learning rate across all trials.

\section{Action selection}
As with learning, our model selects actions (motor commands) based on context inference. If the identity of the current context is known, the forward model for this context is used to select the current action. However, if uncertainty over the context exists, the selected action is influenced by all the possible contexts, with a weight directly related to how likely each one of those contexts is.

@@

Our model offers a different explanation for the observed results: the assymetry is due to the existance of the baseline context. As discussed above, action selection in our model is done via a weighted avarage:
\begin{equation}
p(a_t | s_t) \propto \displaystyle \sum_{m \in M} p(m | s_t, s_{t-1}, a_{t-1}) p(a_t | m)
\end{equation}
In experiments without cued contexts, the baseline model $m_O$ has a non-zero probability $p(m | s_t ...)$. When a new block of trials starts (e.g. in the transition from 3A to A), a switch is inferred by the model (given feedback after the first trial) and $m_O$ becomes more likely (given that $m_{3A}$ has been ruled out. Therefore, in these first trials, action selection has a component guided by the baseline model, in which no extra compensatory force is applied, essentially ``pulling'' adaptation towards zero (no compensatory force). In the first group, this initial pull towards zero accelerates the transition towards A because $3A > A > 0$, but on the second group, it slows down the switch because $A > 0 > -A$.

@@

\begin{figure}
\centering
\includegraphics{./figures/figure_3.png}
\caption{Motor error when switching back to a previously learned adaptation. Following the block structure from \cite{Davidson_Scaling_2004}, the first 40 trials of the switch from +8 (yellow) or -8 (blue) to 4 is shown. (A) Simulations from our model, with the task parameters from \cite{Davidson_Scaling_2004}. On the y axis, the error is shown in arbitrary units, related to centimiters through a monotonically increasing transformation with the same origin. (B) Data adapted from \cite{Davidson_Scaling_2004}, showing the error in the same trials as (A).}
\label{fig:davidson-2004}
\end{figure}

@@

\cite{Vaswani_Decay_2013} explored in detail human behavior during an error-clamp block in a shooting movement paradigm with a mechanical arm. The authors found that during an E block at the end of each experiment, there was a lag of a few trials (depending on participant) before their motor behavior changed from that of the previous block. After that, the exerted force slowly dropped towards zero throughout tens of trials, but never reaching values around zero. Participants were divided into four groups, each of which going through a different block sequence: (1) A-E, (2) O-A-E, (3) (-A/2)-A-E, and (4) (-A)-E. No pauses were made during the experiment nor were there any contextual cues, so transitions between blocks were not signaled to participants. However, the mechanical arm used in the experiment provided force-feedback to participants.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./figures/figure_4.png}
\caption{Adaptation during error clamp trials. (A) Simulated adaptation during the error-clamp trials for the three groups of participants in \cite{Vaswani_Decay_2013}, using the same colors. The solid line is the average across 10 runs (i.e. 10 simulated participants) and the shaded area represents the standard deviation. The vertical dashed line is the start of the error-clamp trials.(B) Corresponding experimental data adapted from \cite{Vaswani_Decay_2013}. (C) Inference over the current context, where contexts are color coded: black for baseline, orange for the counter-clockwise force, purple for the clockwise force and brown for clockwise force with half strenght. The lines represent the posterior probability of each ocontext at every trial, while the background color represents the real context. An olive-colored background represents error-clamp trials. As in (A), solid lines represent the average across all runs and shaded areas represent the standard deviation. (D) Visualization of the lag before error-clamp trials are detected. Each line represents one run (groups 1.1, 1.2 and 1.3, 10 runs per group).}
\label{fig:vaswani-2013}
\end{figure}

This effect is explained in our model in terms of context inference. In \fref{fig:vaswani-2013}C, the inference over context is shown for each group separately. Context inference works reliably until the error-clamp trials start, which do not correspond to any of the known contexts. This causes the agent to infer the combination of some of the known contexts that best fits the observations. Depending on the contexts previously learned by the agent (which change from group to group), inference during E trials will vary: groups 1.1 and 1.4 have exactly the same behavior, where the previous context (A and -A, respectively) slowly dwindles. These agents will slowly lower the force applied until eventually reaching zero (after hundreds of trials; not shown). In contrast, group 1.3 has learned the additional -A/2 context (represented by brown in \fref{fig:vaswani-2013}C), which has a non-zero posterior probility during E trials, pushing the agent's adaptation force more quickly towards zero. Group 1.2 behaves similarly to 1.1, with the exception that the baseline context, which was recently seen, plays a bigger role during E trials, making the agent reduce its force during E trials slightly more slowly than groups 1.1 and 1.4.

Our model also reproduces the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. To show this, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context (A for groups 1.1, 1.2 and 1.3, -A for 1.4) for each simulated run (one line per run, 10 runs per group, all groups). It can be seen that the moment in which the drop starts, as well as the speed of the drops, depends on the run.

@@

\subsection{Extended explanatory power}
Context is often directly observable, either via environmental cues or via feedback observed after actions. In many behavioral experiments, for example, the context of the current block of trials is easily identifiable by the color of a cue light \citep[e.g.][]{Ethier_Spontaneous_2008}, the posture of the hand \citep[e.g.][]{Gandolfo_Motor_1996}, the target of the movement \citep[e.g.][]{Lee_Dual_2009} or by the workspace in which the task is being performed \citep[e.g.][]{Shadmehr_Adaptive_1994}. The context can also be selected by the agent itself. For example, a person can decide the level of care devoted to moving an egg or a lead ball around, which determines the way in which the person would move their hands. In these cases, the context is clearly determined and known to the agent and switches between different adaptations are immediate.

@@

One of the strenghts of this model is that its general principle does not
depend on the specifics of the experiment to which it is applied. However, for
clarity, we will connect the different parts of the model to different
experiments.

Before proceding with the model description, let us introduce nomenclature:
\begin{itemize}
\item $t$: Time at the moment an observation is made, before inference is done.
\item $\theta_t$: Observation. Note that this is assumed to be a noisy
observation of the generalized state (see below) given by $\theta_t = h(z_t)$.
\item $s_t$: Hidden state, e.g. the position of the hand in reaching
experiments. It does not depend on the context.
\item $a_t$: Action taken after observing $\theta_t$ and inferring the context.
\item $c_t$: Context inferred after observing $\theta_t$.
\item $\omega_t$: Contextual information. This term includes, e.g. visual cues,
proprioceptive feedback, performance feedback.
\item $z_t$: Generalized state, $z_t = \{s_t, c_t, \omega_t\}$. Observed noisily
as $\theta_t$.
\end{itemize}

Context inference can work through context-specific (possibly unreliable) sensory cues,
or through the dynamics of the system (e.g. expected outputs of motor
commands), using prediction error as the force behind inference. In this case,
the generative model can is written as follows:
\[
p(z_t, z_{t-1}, a_{t-1}, \theta_t, \theta_{t-1}) = p(\theta_t | z_t)p(z_t|\theta_{t-1}, a_{t-1})
\]
and inferring the current (hidden) generalized state is done with Bayes' theorem:
\[
  q(z_t | \theta_t, \theta_{t-1}, a_{t-1}) \propto p(\theta_t | z_t)p(z_t|\theta_{t-1}, a_{t-1})
\]

Additionally, after having inferred the current state, motor adaptation is given by
\[
q(\gamma | z_t, z_{t-1}, a_{t-1}) \propto p(z_t | z_{t-1}, a_{t-1}, \gamma)p(z_{t-1})p(\gamma)
\]

where $\gamma$ are the parameters of the internal model for the dynamics,
i.e. the function $f: (z_t, a_t) \rightarrow z_{t+1}$, which determines the beliefs of
the agent regarding how the system evolves after having taken action $a_t$.

Having inferred the generalized state (including context), a decision is made
sampling from a distribution:
\begin{equation}
p(a_t | z_t, \beta) = \displaystyle\sum_{m \in M}q(C_t)p(a_t | z_t, \beta, m)
\end{equation}
where $M$ is the set of all relevant forward models (see below), and
$p(a_t | ... m)$ is the probability distribution over actions prescribed by the
forward model $m$.

>>>>

While these general models of adaptation explain the most common phenomena observed in experiments, a number of known phenomena remain outside of their scope. For example, it is known that adaptation rate is reduced in situations where the environment is unstable and unpredictable \citep{Herzfeld_memory_2014}, or situations in which errors are small \citep{Marko_Sensitivity_2012} and slowly introduced \citep{Huang_Persistence_2009}. Learning new adaptations has also been found to depend on the history of adaptations learned \citep{Vaswani_Decay_2013}.

@@

\subsection{The model}
The model we present has three main components: (1) context-inference, (2) motor adaptation and (3) action selection. Each of these components informs the ones that follow: motor adaptation is informed by context inference, and action selection is informed directly by both motor adaptation and by context inference. A schematic representation of the model can be seen in \fref{fig:model}.

\begin{figure}
\centering
\includegraphics[]{./figures/figure_1.png}
\caption{Description of the model. (A) Definition of a context in terms of its components. (B) Update equations for context inference, motor adaptation and action selection. $\zeta$ represents the identity of the context as a categorical variable, $\vartheta$ are observations, $c_t$ are motor commands at trial $t$, $\phi$ are the parameters of the forward model and $s$ are the hidden states of the environment (e.g. the position of the hand). Note that because action selection is not performed via Bayes' theorem, the notation $p(c_t)$ is used. (C) Schematic representation of the model. At trial $t$, an observation $\vartheta_t$ is received, from which the current context is inferred ($q(\zeta)$), which informs motor adaptation ($q(\phi)$), which informs action selection ($p(c)$). After an action is selected, a new observation $\vartheta_{t+1}$ is received and the process repeated. Note that the posteriors over $\zeta$ and $\phi$ are used as priors at the next trial.}
\label{fig:model}
\end{figure}

We now briefly describe the three components of the model separately. For clarity, we start with the second component (i.e. motor adaptation), in order to introduce the terminology that we use throughout this work. To see how these three components affect each other, see \fref{fig:model}B and C.

@@

\subsection{Context inference}
In our model, the idea of context is quite general (see \fref{fig:model}A). The context comprises the relevant elements of the environment (e.g. the strength and direction of the wind when walking), of the task at hand (immediate and long-term goals, reward structure, etc) and task-related information internal to the decision-making agent (e.g. forward model, decision-making algorithms, motor adaptation equations, expected observations). Previous models \cite[e.g.][]{Wolpert_Multiple_1998,Imamizu_Neural_2008,Oh_Minimizing_2019} have focused on the elements of the context which are directly related to the internal models: given a past action, the different internal models (e.g. two forward models $f_B$ and $f_H$) make different predictions for the outcome of that action; the model that best predicts that action is deemed to be the ``correct'' model, i.e. the model that best represents the context.

@@

\subsection{Action selection}
Action selection is affected by the current context via the context-dependent forward model that is used. In our model, action selection is made by building a distribution over available actions which is a weighted sum of the distributions given by the existing forward models, where the weights are obtained by the context inference component of the model. In other words, a distribution is created as follows:
\begin{equation}
p(c_t) \propto \displaystyle\sum_{\zeta \in \Phi}q(\zeta | \vartheta_t, \vartheta_{t-1}, c_{t-1}) p(c_t | \zeta) \label{eq:action-selection}
\end{equation}
where $p(c_t | \zeta)$ corresponds to \eref{eqn:forward-model}. From this distribution, the current action (motor command $c_t$) can be sampled and carried out.

\section{Experimental results}
In this section, we go through a number of experimentally-observed phenomena and show that context inference, as done in our model, provides a unifying explanation for all of them.

For clarity, we first introduce necessary terminology. As an example, we will use a typical motor adaptation task in which participants have to make forward-backward reaching movements while holding the handle of a mechanical arm that exerts a force. Depending on the trial, the arm might exert a curl force in a clockwise or counter-clockwise direction, or no force at all. Let us define the baseline context O as that in which the mechanical arm exerts no force. Context A can be defined as that in which the arm exerts the clockwise curl force and context B counter-clockwise. Additionally, abusing notation, we can say that $B = (-A)$, as the forces have the same magnitude but point in opposite directions. We can also talk of contexts such as A and A/2, which means that both have the same direction of the adaptation (e.g. clockwise), but the second one has half the magnitude. Finally, error-clamp trials, in which the mechanical arm forces the participant to make straight-line movements, are represented with the letter E.

@@

\subsection{Savings and slow/fast switching}
Savings refers to the ability of humans and other animals to remember a previously-learned adaptation and apply it without having to re-learn it. Savings is almost universally observed in human participants \cite{Brashers-Krug_Consolidation_1996,Shadmehr_Functional_1997,Medina_Mechanism_2001,Smith_Interacting_2006,Zarahn_Explaining_2008}. In an O-A-O-A experiment, for example, savings would be seen in the second A block in the form of a much higher adaptation rate higher than that observed during the first A block.

In this section, we discuss savings, and the related phenomenon of deadaptation, in terms of the switching between forward models. This has been modeled before \citep[e.g.][]{Wolpert_Multiple_1998,Oh_Minimizing_2019}, but we show that expanding forward model switching into context inference explains facets of savings that these previous models do not.

@@

To our knowledge, there are no experiments in which no contextual information is available. However, many experiments \citep[e.g.][]{Huang_Persistence_2009,Brennan_Decay_2015,Smith_Interacting_2006} contain blocks of error-clamp trials in which participants are not provided with contextual information. These blocks could be considered uncued, however, as we discuss in the following sections, contextual information is still available to participants. In this section, we focus on cued and partially-cued experiments and leave error-clamp trials to the following sections.

@@

\begin{figure}
\centering
\includegraphics{./figures/figure_2.png}
\caption{Savings and de-adaptation. Data from our simulations (left column) compared to data from \cite{Kim_Neural_2015} and \cite{Oh_Minimizing_2019} (right column). In all panels, the blue line represents the adaptation displayed by the agent as a function of trial number. The black lines represent the optimal adaptation, i.e. the size of the visuomotor rotation during the task. (A) Experiment by \cite{Kim_Neural_2015} with two rotations (-40 and 40 degrees), in addition to baseline. The switches between contexts were visually cued and therefore switches are immediate, as can be seen in both simulations (left) and the experimental data from \cite{Kim_Neural_2015} (right). (B) An experiment in which the agent must adapt to a 20 degree visuomotor rotation during shooting movements. It can be seen in both the simulations (left) and experimental data from \cite{Oh_Minimizing_2019} (right) that after context switches (e.g. at trial 120), the agent returns to the adaptation of the previous block, with one or two trials of delay. (C) Same experiment as (B) but  with a 10 degree adaptation. In this case, savings are much slower.} \label{fig:oh-2019}
\end{figure}

\cite{Oh_Minimizing_2019} performed a similar experiment in which contexts were not visually cued. In addition, the authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. Participants in \cite{Oh_Minimizing_2019} showed immediate switching in the first experiment, while switching was much slower in the second experiment. As can be seen on the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large, and slow, noisy switches when it is low. This difference is explained by our model in terms of the size of the adaptation: as the size of the adaptation is lower (10 degrees), it becomes more difficult to distinguish errors made by incorrectly identifying the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

@@

\section{Slower adaptation in volatile and unpredictable environments}
In our model, adaptation is gated by the uncertainty on the current context: a prediction error observed at trial $t$ will effect motor adaptation with a magnitude proportional to the size of the error, where the proportionality constant is related to the uncertainty over the context in which the error was observed: the higher the uncertainty, the lower the size of the adaptation.

The most direct evidence for the effects of context inference on the rate of adaptation come from the experiments by \cite{Herzfeld_memory_2014}. The authors showed that the volatility of the environment, expressed in terms of unpredictable, stochastic transitions between A and -A (and vice versa), change the speed at which adaptation occurs.

\cite{Herzfeld_memory_2014} used an experimental paradigm in which the force that a mechanical arm exerts on the participant's hand can change from trial to trial between O, A and -A. They divided participants into three groups for which the context transition from one trial to the next had differing probabilities, depending on the group: for one group transitions were very uncommon, for the second group they were somewhat common and for the third group they were almost every trial. Importantly, these transitions were uncued, and the participant could only infer that a transition had occurred after a motor error was observed.

@@

\begin{figure}
\centering
\includegraphics{./figures/figure_3.png}
\caption{Learning under unstable and unpredictable contexts. 30 trials were simulated in the task defined by \cite{Herzfeld_memory_2014}, repeated 300 times for better statistics. Two conditions were used, following the naming convention from \cite{Herzfeld_memory_2014}: $z=0.5$ is an unpredictable condition, where the context has a 0.5 probability of switching from trial to trial. $z=0.9$ is a stable condition, where the context has a 0.9 probability of staying the same from trial to trial. (A) Adaptation as a function of trial number, averaged across all runs. (B) Error sensitivity, binned into bins of 5 trials, averaged across all runs. The error is in arbitrary units, causing sensitivity to error (y axis) to also be in arbitrary units. Shaded areas are SEM.}
\label{fig:herzfeld-2014}
\end{figure}

To account for volatility, our model includes a term in context inference which assumes a transition can occur from the previous context to all other possible contexts (see Methods). Effectively, this transition term gives the model uncertainty on the current context, regardless of observations. The higher the volatility, the higher the uncertainty. This higher uncertainty is what lowers the observed learning rate across all trials.

\section{Uncertainty over contexts affects action selection}
As with learning, our model selects actions (motor commands) based on context inference. If the identity of the current context is known, the forward model for this context is used to select the current action. However, if uncertainty over the context exists, the selected action is influenced by all the possible contexts, with a weight directly related to how likely each one of those contexts is.

@@

Our model offers a different explanation for the observed results: the assymetry is due to the existance of the baseline context. As discussed above, action selection in our model is done via a weighted avarage (see \eref{eq:action-selection}). In experiments without cued contexts, the baseline model $m_O$ has a non-zero probability $p(m | s_t ...)$. When a new block of trials starts (e.g. in the transition from 3A to A), a switch is inferred by the model (given feedback after the first trial) and $m_O$ becomes more likely (given that $m_{3A}$ has been ruled out. Therefore, in these first trials, action selection has a component guided by the baseline model, in which no extra compensatory force is applied, effectively ``pulling'' adaptation towards zero (no compensatory force). In the first group, this initial pull towards zero accelerates the transition towards A because $3A > A > 0$, but on the second group, it slows down the switch because $A > 0 > -A$.

@@

\begin{figure}
\centering
\includegraphics{./figures/figure_4.png}
\caption{Motor error when switching back to a previously learned adaptation. Following the block structure from \cite{Davidson_Scaling_2004}, the first 40 trials of the switch from +8 (yellow) or -8 (blue) to 4 is shown. (A) Simulations from our model, with the task parameters from \cite{Davidson_Scaling_2004}. On the y axis, the error is shown in arbitrary units, related to centimiters through a monotonically increasing transformation with the same origin. (B) Data adapted from \cite{Davidson_Scaling_2004}, showing the error in the same trials as (A).}
\label{fig:davidson-2004}
\end{figure}

@@

\cite{Vaswani_Decay_2013} explored in detail human behavior during an error-clamp block in a shooting movement paradigm with a mechanical arm. The authors found that during an E block at the end of each experiment, there was a lag of a few trials (depending on participant) before their motor behavior changed from that of the previous block. After that, the exerted force slowly dropped towards zero throughout tens of trials, but never reaching values around zero. Participants were divided into four groups, each of which going through a different block sequence: (1) A-E, (2) O-A-E, (3) (-A/2)-A-E, and (4) (-A)-E. No pauses were made during the experiment nor were there any contextual cues, so transitions between blocks were not signaled to participants.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./figures/figure_5.png}
\caption{Adaptation during error clamp trials. (A) Simulated adaptation during the error-clamp trials for the three groups of participants in \cite{Vaswani_Decay_2013}, using the same colors. The solid line is the average across 10 runs (i.e. 10 simulated participants) and the shaded area represents the standard deviation. The vertical dashed line is the start of the error-clamp trials.(B) Corresponding experimental data adapted from \cite{Vaswani_Decay_2013}. (C) Inference over the current context, where contexts are color coded: black for baseline, orange for the counter-clockwise force, purple for the clockwise force and brown for clockwise force with half strenght. The lines represent the posterior probability of each ocontext at every trial, while the background color represents the real context. An olive-colored background represents error-clamp trials. As in (A), solid lines represent the average across all runs and shaded areas represent the standard deviation. (D) Visualization of the lag before error-clamp trials are detected. Each line represents one run (groups 1.1, 1.2 and 1.3, 10 runs per group).}
\label{fig:vaswani-2013}
\end{figure}

This effect is explained in our model in terms of context inference. In \fref{fig:vaswani-2013}C, the inference over context is shown for each group separately. Context inference works reliably until the error-clamp trials start, which do not correspond to any of the known contexts. This causes the agent to infer the combination of some of the known contexts that best fits the observations. Depending on the contexts previously learned by the agent (which change from group to group), inference during E trials will vary: groups 1.1 and 1.4 have exactly the same behavior, where the previous context (A and -A, respectively) slowly dwindles. These agents will slowly lower the force appl ied until eventually reaching zero (after hundreds of trials; not shown). In contrast, group 1.3 has learned the additional -A/2 context (represented by brown in \fref{fig:vaswani-2013}C), which has a non-zero posterior probility during E trials, pushing the agent's adaptation force more quickly towards zero. Group 1.2 behaves similarly to 1.1, with the exception that the baseline context, which was recently seen, plays a bigger role during E trials, making the agent reduce its force during E trials slightly more slowly than groups 1.1 and 1.4.

Our model also reproduces the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. To show this, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context (A for groups 1.1, 1.2 and 1.3, -A for 1.4) for each simulated run (one line per run, 10 runs per group, all groups). It can be seen that the moment in which the drop starts, as well as the speed of the drops, depends on the run, reproducing the different lags observed in experimental data. This can also be seen directly in context inference, displayed as the shaded areas in \fref{fig:vaswani-2013}C.

@@

\subsection{Alternative explanations}
For each of the representative studies presented in the Results section, the authors put forward explanations for the phenomena they observed in their data. While our model provides a single explanation for all these phenomena, it is not necessarily at odds with those presented by the respective authors and might instead act on top of the explanations the authors presented.

\cite{Oh_Minimizing_2019} posited that the difference between their two groups (~\fref{fig:oh-2019}B and C) can be explained by considering that the smaller visual rotation of 10 degrees caused participants to update their baseline forward model instead of creating a new forward model for the context. If such an account were true, it would predict that when participants walk away from the experiment they would have trouble adapting to the no-rotation reality, having to re-learn it. Our model instead would suggest that participants would immediatly switch back to their baseline movements the moment they walk away from the computer screen, as that is a clear indication that the context has switched back to baseline. More generally, our model predicts that the observed difference between the two groups would disappear if clear sensory contextual cues were added to the experiment. While we believe our model provides a better explanation for the observations by \cite{Oh_Minimizing_2019}, the effect the authors describe has been documented before in chimpanzees \citep{Kojima_Memory_2004}; however, to our knowledge such an effect has not been observed in humans in tasks like that by \cite{Oh_Minimizing_2019}.

The paradigm by \cite{Herzfeld_memory_2014} was designed to test the hypothesis that a memory of errors affected the speed at which participants adapted to curl forces. While it has been shown that sensitivity to error can be modulated \cite{Marko_Sensitivity_2012}, and indeed in our model it depends on the amount of experience (see \sref{sec:interpreting-hyperparameters}, to our knowledge there are no other experiments suggesting that it depends on the history of errors. As before, the introduction of clear sensory contextual cues would allow us to gauge the importance of context inference and of previous errors in the observed data. The same can be said of the experiments by \cite{Davidson_Scaling_2004}, in which different contexts involve movement in different directions, which requires the use of different muscles. This alone could cause delays in switching unrelated to context inference [CITATION NEEDED on the delay in switching muscles?], but we expect the effects to be insufficient to explain the observed data.

\subsection{Further experimental evidence}
Context is often directly observable, either via environmental cues or via feedback observed after actions. In many behavioral experiments, for example, the context of the current block of trials is easily identifiable by the color of a cue light \citep[e.g.][]{Ethier_Spontaneous_2008}, the posture of the hand \citep[e.g.][]{Gandolfo_Motor_1996}, the target of the movement \citep[e.g.][]{Lee_Dual_2009} or by the workspace in which the task is being performed \citep[e.g.][]{Shadmehr_Adaptive_1994}. The context can also be selected by the agent itself. For example, a person can decide the level of care devoted to moving an egg or a lead ball around, which determines the way in which the person would move their hands. In these cases, the context is clearly determined and known to the agent and switches between different adaptations are immediate.

@@

Before proceding with the model description, let us introduce nomenclature:
\begin{itemize}
\item $t$: Time at the moment an observation is made, before inference is done.
\item $\vartheta_t$: Observation. Note that this is assumed to be a noisy
observation of the generalized state (see below) given by $\vartheta_t = h(z_t)$.
\item $s_t$: Hidden state, e.g. the position of the hand in reaching
experiments. It does not depend on the context.
\item $c_t$: Motor command (action) taken after observing $\vartheta_t$ and inferring the context.
\item $\zeta_t$: Context inferred after observing $\vartheta_t$.
\item $\omega_t$: Contextual information. This term includes, e.g. visual cues,
proprioceptive feedback, performance feedback.
\item $z_t$: Generalized state, $z_t = \{s_t, \zeta_t, \omega_t\}$. Observed noisily
as $\vartheta_t$.
\end{itemize}

Context inference can work through context-specific (possibly unreliable) sensory cues,
or through the dynamics of the system (e.g. expected outputs of motor
commands), using prediction error as the force behind inference. In this case,
the generative model can is written as follows:
\[
p(z_t, z_{t-1}, c_{t-1}, \vartheta_t, \vartheta_{t-1}) = p(\vartheta_t | z_t)p(z_t|\vartheta_{t-1}, c_{t-1})
\]
and inferring the current (hidden) generalized state is done with Bayes' theorem:
\[
  q(z_t | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | z_t)p(z_t|\vartheta_{t-1}, c_{t-1})
\]

Additionally, after having inferred the current state, motor adaptation is given by
\[
q(\gamma | z_t, z_{t-1}, c_{t-1}) \propto p(z_t | z_{t-1}, c_{t-1}, \phi)p(z_{t-1})p(\phi)
\]

where $\phi$ are the parameters of the internal model for the dynamics,
i.e. the function $f: (z_t, c_t) \rightarrow z_{t+1}$, which determines the beliefs of
the agent regarding how the system evolves after having taken action $c_t$.

Having inferred the generalized state (including context), a motor command is
sampled from a distribution:
\begin{equation}
p(c_t | z_t) = \displaystyle\sum_{m \in M}q(\zeta_t)p(c_t | z_t, m)
\end{equation}
where $M$ is the set of all relevant forward models (see below), and
$p(a_t | ... m)$ is the probability distribution over actions prescribed by the
forward model $m$.
